---
title: "Morality and Religious Belief"
author: "Ian Myers"
date: "3/17/2020"
geometry: margin=4cm
output:
  pdf_document: default
  html_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width= 3, fig.height= 3 )
```
 
## Question
 
Of practical questions, some are moral, and some are non-moral. It should suffice to distinguish the two by remarking that while, in the case of the non-moral, the standard against which actions are properly judged is the benefit of an individual, in the case of the moral, the standard is something other than, higher than, the benefit of an individual. As manifold as these higher standards may be, they are united in drawing our eyes away from that concern that might appear to be deepest in us, the concern for our own well-being. In light of this distinction, we can understand different individuals to be at different locations on a continuum of how salient moral standards are to their evaluations of answers to practical questions. Those who are more “moral” will understand moral standards to be more important for evaluating answers to more practical questions, and those who are less “moral” will understand moral standards to be less important and relevant in fewer matters. The question I here investigate is whether an individual’s movement on this continuum from more moral to less moral predicts a negative change in that individual’s certainty in the existence of a god.
 
## Literature
 
Before stating my hypothesis and the theoretical grounds for it, I will briefly indicate how the state of the literature on the causal relations between political identity, religious identity, and relative inclination to different “moral foundations” suggests both the necessity of raising this question and guidelines for how to understand it. With some exceptions, scholars prior to the early 2000’s primarily understood religious belief and identification as a basic fact, i.e. as a fact which explains but is not explained by variables in the domain ruled by political science. According to this understanding, the increasing alignment of Republican vs. Democrat identification with religious vs religious none identification is a consequence of individuals sorting themselves politically in accord with the prior fact of their religious identification (see, e.g., Campbell et al. 1960; Layman 2001). This understanding, however, has been questioned and to a large extent refuted in the past 20 years as scholars have found that the causality works as well in the other direction, from political identification to religious identification and belief (Hout and Fischer 2002; Patrikios 2008; Baker and Smith 2009; Putnam and Campbell 2010). But much of the work clarifying this new causal direction has been limited to clarifying the effect of political identification on religious identification and congregational affiliation, leaving open the question of how religious beliefs themselves are affected (Patrikios 2008; Djupe et al. 2017; see especially Hout and Fischer 2002, which notes that those who change to religious nones do not always appear to change their beliefs). And those that have looked at religious belief as an effect of political identification (e.g. Campbell et al. 2018) have largely ignored what, it seems to me, might be the deeper cause. 

What is perhaps the most common (see, e.g., Djupe et al. 2017) theoretical explanation of the power of political identification to shape religious identification and belief can be put, however crudely, in the following way: individuals tend to sort themselves politically and socially in consistent ways; the Republican party has become popularly associated with religiosity; religious Democrats will therefore be likely to resort themselves either politically or religiously; political sorting is more or as fundamental as religious sorting; religious Democrats will therefore be likely to resort themselves as religious nones, and some might even take the next step of modifying or dropping their religious beliefs. Such a theory is no doubt sensible and, more than that, likely correct, but it leaves open the question of whether there is an additional, deeper reason (deeper than the what might be understood as accidental fact of a common association of religiosity with the Republican party) for the effect political beliefs have on religious beliefs. My hypothesis is that it is not only an individual’s political identification or political beliefs, but also their evaluative framework (morality vs non-morality), which both causes and is caused by their particular political beliefs, that causes their religious beliefs. 
 
## Theory
	
The concept that most needs to be explained here is that of the moral individual, or of the moral continuum which individuals fall at different points on, which I will call the continuum of morality. It will be helpful first to state what this concept does not signify. Morality as I am using it here does not signify a genetic predisposition to evaluate answers to practical questions with reference to certain standards rather than others. Nor does it signify a manner of thinking which simply explains or causes the particular value judgments that an individual makes. According to that view, when we are faced with a practical question, we come to answers by inferring practical propositions from the axioms that our moral standards constitute. This, however, is of course not how even the most thoughtful of us make most of our judgments when faced with practical questions. What happens for the most part, in the theory I am supposing, is that when faced with practical questions we have a gut reaction that carries along with it an answer to that question (an opinion). That gut reaction and the opinion that comes along with it arises prior to thoughtful reflection on the question, but it in most cases implies certain more general opinions about the facts of practical human life. Someone who opines that abortion should be illegal after the second term, for example, likely can be led to say that fetuses after the second term are humans, that humans have a right to life, and that one has an obligation to respect the rights of others. 

Because we have a great number of practical opinions, we are liable to have opinions implied by those opinions that contradict each other. When such a contradiction first becomes apparent, and when no psychological obstacle hinders us from recognizing the full implications of that contradiction, we discard that one of the contradictory opinions which grips us less strongly. On the other hand, as the consistency of our opinions slowly increases, we begin to see more and more what the broader, more fundamental opinions implied by those opinions are. These more fundamental practical beliefs or opinions are in some way the causes of our particular practical opinions, but, again, we are not aware of them or this causal power of theirs until we investigate what is implied by our practical opinions. 

Now there is a hypothesis common to the ancient political philosophers that belief in a god or gods is bound up with belief in the importance of justice, or of the importance of abiding by standards of conduct which are in tension with and higher than the principle of acting in one’s own interest. There follows from that hypothesis the testable hypothesis that the genuine rejection of the belief in the importance of justice and the replacement of that belief with the belief that one’s enlightened interest is the right principle for action leads to the weakening and eventual disappearance of the belief in a god or gods. The theoretical basis for that testable hypothesis is something as follows. There is a tension in the common understanding of justice. On the one hand, we say that justice is difficult and often involves the sacrifice of our good, and our anger at those who get away with injustice indicate that we believe that justice is in an important way bad and injustice good. On the other hand, parents, who above all want what is best for their children, exhort their children to be just. Moreover, only the wicked would choose to be unjust, and surely no one would say that wickedness is good. According to the common understanding of justice then, justice is both good and bad. This tension in the common understanding is caused by our understanding just actions, and especially of the most spectacular just actions, to be noble sacrifices that are somehow good precisely on account of being noble sacrifices. It is of course difficult to explain how something can be good precisely on account of its being a sacrifice (i.e. on account of its being bad for us). What then is the explanation given by the ancient political philosophers for this perplexing phenomenon, which consists, on the one hand, in the nearly universal human agreement in the compatibility of one’s good and a sacrifice of one’s good and, on the other hand, in the just as universal human wavering on the question of the goodness for justice? The clear equation of the sacrifice of our own good and our own highest good is untenable for human minds on account of its being a direct contradiction, but one of the important ways in which such a contradictory equation could be made psychologically maintainable is by the addition of a secret hope that, since we understand sacrifices to be grounds for desert, our justice will make us deserving of great rewards. That hope must remain secret, and the contradictory equation must remain in some form maintained because were they not, our choices to act justly would be transparently motivated by our concern for our own good—were we to say that justice is not good for its own sake but good because it will lead to our being rewarded, we would undermine the basis for our being rewarded, which we understand to be our willingness to sacrifice our own good without expecting a reward. But just human beings are not always or even as a rule rewarded for their sacrifices. For this reason, a just human being is more likely to hope, and hence even come to believe, that a god that rewards the righteous and punishes the wicked exists, and a human being less concerned with justice is less likely to hope for and believe in such a god. In short, according to the ancient political philosophers, one of the great causes of belief in a god or gods is our longing for a being or force, not supplied by the natural order, which will reward justice and punish injustice. And this proposition can be generalized such that by ‘justice’ is meant all actions and dispositions to action which look not to the standard of one’s own good but to another standard understood to be higher. We can thus express the proposition in its generalized form as follows: the greater the saliency of moral standards to an individual, the more likely they are to believe in a god.

More detailed and precise statements of this thought, along with discussions of its implications, can be found in the chapters "Theages" and "Euthyphro" of Bruell 1999 and pp. 69-70 and 92-100 of Leibowitz 2010.
 
## Research Design
 
The unit of analysis here is the individual, and the data used comes from the 2012 Portraits of American Life Study (PALS). This data is two-wave panel data collected by phone in 2006 and 2012 from U.S. civilians aged 18 or over living in the U.S.

I estimate two models, a linear model and a gologit model. The dependent variable, change in an individual's certainty about the existence of god, is operationalized in two ways (for the two models). In both waves of the survey, participants were asked to say whether they strongly agree, somewhat agree, etc. to the statement “I definitely believe in God.” For my linear model, I count strongly agree as 4 and strongly disagree as 0 and subtract the second wave number from the first wave number. This will result in a variable with 9 possible values ranging from -4 to 4. For my gologit model, I operationalize change in certainty about the existence of god by reducing the variable to one with three values: -1 for a decrease from wave 1 to 2 in agreement with the statement; 0 for no change; and 1 for an increase in agreement with the statement. 

The independent variable, the salience of moral standards in the matter of political questions, is operationalized as follows. Both waves of the survey contain two questions in which participants are asked to say whether a matter is always morally wrong, usually morally wrong, sometimes morally wrong, never morally wrong, or not a moral issue. The first of these questions concerns “having an abortion when the fetus is old enough to survive on its own outside the mothers womb,” and the second concerns “using genetic engineering, that is changing a persons DNA or genes, to create a baby that is smarter, stronger, or better looking.” I assign to the response “always morally wrong” the value 4 and to the response “not a moral issue” the value 0. I then subtract the individual’s wave 1 responses from their wave 2 responses. There are in this way two main independent variables with values ranging from -4 to 4, where positive values indicate an increase for the individual in the salience of moral standards to the question asked. 

The great limitation of this operationalization of the independent variable of interest lies in the questions the participants are asked about. That the questions are only two in number and both relate to the narrow matter of human reproduction (abortion and genetic engineering) means that this variable cannot be understood to signify the salience for the individual of moral standards to practical and political questions generally—one can imagine an individual who thinks neither abortion nor genetic engineering is a moral issue but who has strong moral convictions about other matters. With that being said however, since the salience for an individual of moral standards to one set of questions would seemingly predict with some accuracy the salience for them of moral standards to practical and political questions in general, the theory detailed above should still support the modified hypothesis that individuals whose estimation of reproductive issues as moral issues decreases will be more likely to decrease in certainty about the existence of a god. 

The linear and gologit model I estimate are as follows. The control variables used in the linear model are the same as those listed in the Gologit model results below.

### Linear Model

$$\Delta Certainty=\beta_{0}+\beta_{1}\Delta AMoralSaliency +\beta_{2}\Delta GEMoralSaliency+\beta_{3:n}Controls+e$$

### Gologit Model

$$Pr(Y_{i}>j)=\frac{exp(X_{i}'\beta_{j} -\mu_{j})}{1+exp(X_{i}'\beta_{j}-\mu_{j})}, j=1,2,...,m-1$$
where $X_{i}'\beta_{j}$ is the same as the right side of the linear model without the error random variable or intercept parameter. Which $\beta$ parameters are restricted and which unrestricted can be seen in the Gologit model results below.

## Results (Linear Model)

The estimated parameters for the linear model that are significant are presented below in table 1. While change in an individual's take on the salience of moral standards to the question of abortion did not have a statistically significant effect on their change in certainty about the existence of God, change in an individual's take on the salience of moral standards to the question of genetic engineering of humans did. This indicates that changes in at least some moral beliefs might be causally related to changes in certainty in the existence of God independently of PID or party affiliation. 

The graph below shows the expected change in certainty in the existence of God given a change in the salience of morality to questions of genetic engineering for an individual with mean values for the other observed variables. To interpret this graph, one has to assume that certainty is a continuous variable that was artificially discretized by the survey questions. 

```{r, include=FALSE}
library(MASS)
library(brant)
library(haven)
library (dplyr)
library(stargazer)
library(ggplot2)

#Importing Data and removing responses to Moral and Belief Questions like "I don't know"

data <- read_dta("PortraitsData.DTA")

data <- subset(data,MA_1<6 & MA_1>0 & MA_1W1<6 & MA_1W1>0 & MA_4<6 &MA_4>0 & MA_4W1<6 & MA_4W1>0 & RA_1W1 >0 & PO_1W1 > 0 & PO_1W1 < 6 & PO_1 < 6 & PO_1 > 0)

#Main IV's

data$AbortionDiff <- -(data$MA_1 - data$MA_1W1)
data$GenEngineeringDiff <- -(data$MA_4 - data$MA_4W1)

#Controls

data$PID <- data$PO_1W1
data$PIDDiff <- data$PO_1-data$PO_1W1
data$Income <- data$DMINCW1
data$Age <- data$HRAGEW1
data$Male <- data$HRGENDW1
data$Black <- data$ARE5==2
data$Asian <- data$ARE5==4
data$Hispanic <- data$ARE5==3
data$AmericanIndian <- data$ARE5==5
data$Democrat <- data$PO_6W1==1
data$Republican <- data$PO_6W1==2

#Main DV

data$GodDiff <- -(data$RA_1 - data$RA_1W1)

#Removing NA's

data <- filter_at(data, vars(AbortionDiff,GenEngineeringDiff,Income,Age,Male,Black,Asian,Hispanic,AmericanIndian,Democrat,Republican,GodDiff),
               all_vars(!is.na(.)))

```

```{r pressure, echo=FALSE, include=FALSE}

#linear model and regression

linear <- lm(GodDiff ~ AbortionDiff + GenEngineeringDiff + PIDDiff + PID + Income + Age + Male + Black + Asian + Hispanic + AmericanIndian + Democrat + Republican, data)
stargazer(linear)

```

```{r, echo=FALSE}
#Graphing Change in Certainty in Belief against Change in Saliency of Morality to Genetic Engineering according to Linear Model

average <- cbind(mean(data$AbortionDiff),mean(data$PIDDiff),mean(data$PID),mean(data$Income),mean(data$Age),mean(data$Male),mean(data$Black), mean(data$Asian),mean(data$Hispanic),mean(data$AmericanIndian),mean(data$Democrat),mean(data$Republican))
coefficients <- append(linear$coefficients[4:14], linear$coefficients[2], after = 0)
intercept <- linear$coefficients[1] + (average %*% coefficients)[1,1]
slope <- linear$coefficients[3]

GeneticLinear <- function(x){slope * x + intercept}

# Create the plot
p <- ggplot(data = data.frame(x = 0), mapping = aes(x=x))
p + stat_function(fun = GeneticLinear, color = "red") + 
  xlim(-4,4) + 
  ylim(-4,4) +
  xlab("Change in Saliency of Morality") +
  ylab("Change in Certainty in God")

```

\begin{table}[!htbp] \centering 
  \caption{} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
\cline{2-2} 
\\[-1.8ex] & $\Delta$God Certainty \\ 
\hline \\[-1.8ex] 
 $\Delta$Abortion & 0.041 \\ 
  & (0.026) \\ 
  & \\ 
 $\Delta$Genetic Engineering & 0.076$^{***}$ \\ 
  & (0.023) \\ 
  & \\ 
 $\Delta$PID & 0.057$^{*}$ \\ 
  & (0.033) \\ 
  & \\ 
 Age & 0.006$^{***}$ \\ 
  & (0.002) \\ 
  & \\ 
 Democrat & 0.141$^{*}$ \\ 
  & (0.074) \\ 
  & \\ 
 Constant & $-$0.583$^{***}$ \\ 
  & (0.159) \\ 
  & \\ 
\hline \\[-1.8ex] 
Observations & 727 \\ 
R$^{2}$ & 0.051 \\ 
Adjusted R$^{2}$ & 0.034 \\ 
Residual Std. Error & 0.806 (df = 713) \\ 
F Statistic & 2.941$^{***}$ (df = 13; 713) \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 

## Results (Gologit Model)

The estimated parameters for the gologit model can be seen in table 2 below. *Note*: some coefficients were allowed to vary, while some were not. Which ones were not allowed varied can be determined by looking to the list at the top of the table. Most importantly, neither the coefficients of neither moral difference variable were allowed to vary since both variables satisfied the proportional odds assumption.

As in the estimation of the linear model, the estimate for the abortion morality variable's coefficient is not statistically non-zero whereas the estimate for the genetic engineering morality variable's coefficient is statistically non-zero. This supports, again, the hypothesis that a change in the salience of moral standards to the question of genetically engineering humans for an individual is causally related to a change in that individual's certainty in the existence of God.

Interpreting these coefficients is more difficult than interpreting the coefficients of the linear model. However, since the genetic engineering morality variable satisfies the proportional odds assumption, we can think of its coefficient as analogous to a coefficient in two different logit models, one where the positive response is increase in certainty in the existence of God, and another where the positive respone is decrease in certainty in the existence of God. 

The relation between the coefficient and the probability of these two outcomes is given below, where $x_{i}'\beta_{j}$ refers to the other coefficients and independent variables, and $\beta_{gen}\Delta gen_{i}$ refers to the genetic morality variable and its coefficient.

$$Pr(\Delta Certainty_{i}>0|x_{i}, \Delta gen_{i})=\frac{exp( \beta_{gen}\Delta gen_{i}+x_{i}'\beta_{0} -\mu_{0})}{1+exp( \beta_{gen}\Delta gen_{i}+x_{i}'\beta_{0} -\mu_{0})}$$

$$Pr(\Delta Certainty_{i}\le0|x_{i}, \Delta gen_{i})=1-\frac{exp( \beta_{gen}\Delta gen_{i}+x_{i}'\beta_{-1} -\mu_{-1})}{1+exp( \beta_{gen}\Delta gen_{i}+x_{i}'\beta_{-1} -\mu_{-1})}$$
The graphs of these relationships between the probability of increase and decrease of certainty, on the one hand, and change in the saliency of moral standards to genetic engineering, on the other, with the other variable held constant at their means, are presented below. It can be seen in the first graph that on average, someone who has become completely unsure about the relevance of morality to genetic engineering is very unlikely to have increased in certainty about God while someone who has become certain of the relevance of morality to this has about a 10% chance of becoming more certain of God. On the other hand, it can be seen in the second graph that, on average, someone who has changed to thinking that morality has no place in questions of genetic engineering is more than twice as likely to have become less sure of God's existence as someone who has become certain that genetic engineering is morally wrong. 

For more on interpretation of gologit coefficients in general, and especially on the interpretation of them as odds ratios, see the relevant section of the mathematical appendix.

```{r, echo=FALSE}

#Graphing Probability of Increase in Certainty in Belief against Change in Saliency of Morality to Genetic Engineering according to Gologit Model

average <- cbind(mean(data$AbortionDiff),mean(data$PIDDiff),mean(data$PID),mean(data$Income),mean(data$Age),mean(data$Male),mean(data$Black), mean(data$Asian),mean(data$Hispanic),mean(data$AmericanIndian),mean(data$Democrat),mean(data$Republican))
coefficientsGologit <- c(.12, -.24, -.58, -.03, .01, .3, -.04, -.76, .33, .12, .24, .22)
interceptGologit <- -1.44 + (average %*% coefficientsGologit)[1,1]

GeneticGologit <- function(x){exp(interceptGologit + .17*x)/(1+exp(interceptGologit + .17*x))}

# Create the plot
p <- ggplot(data = data.frame(x = 0), mapping = aes(x=x))
p + stat_function(fun = GeneticGologit, color = "red") + 
  xlim(-4,4) + 
  ylim(0,1) +
  xlab("Change in Saliency of Morality") +
  ylab("Pr of Increased Certainty")


#Graphing Probability of Decrease in Certainty in Belief against Change in Saliency of Morality to Genetic Engineering according to Gologit Model

average <- cbind(mean(data$AbortionDiff),mean(data$PIDDiff),mean(data$PID),mean(data$Income),mean(data$Age),mean(data$Male),mean(data$Black), mean(data$Asian),mean(data$Hispanic),mean(data$AmericanIndian),mean(data$Democrat),mean(data$Republican))
coefficientsGologit2 <- c(.12, -.38, .55, -.03, .01, -.52, -.04, .82, .33, .12, .24, .22)
interceptGologit2 <- -1.44 + (average %*% coefficientsGologit2)[1,1]

GeneticGologitDecrease <- function(x){1-(exp(interceptGologit2 + .17*x)/(1+exp(interceptGologit2 + .17*x)))}

# Create the plot
p <- ggplot(data = data.frame(x = 0), mapping = aes(x=x))
p + stat_function(fun = GeneticGologitDecrease, color = "red") + 
  xlim(-4,4) + 
  ylim(0,1) +
  xlab("Change in Saliency of Morality") +
  ylab("Pr of Decreased Certainty")

```

```{r, echo=FALSE, include=FALSE}

#ologit

#reducing GodDiff to Decrease, Same, and Increase

data$GodDiffSign =sign(data$GodDiff)
data$GodDiffFactor = factor(data$GodDiffSign)

#regression

ologit = polr(GodDiffFactor ~ AbortionDiff + GenEngineeringDiff + Income + Age + Male, data=data, Hess=TRUE)
stargazer(ologit)
brant(ologit)

```

## Conclusion

The results summarized above indicate that changes in an individual's estimation of the saliency of moral standards to certain practical questions are connected to changes in that individual's certainty in the existence of God. Someone who has become less concerned with the moral sides of certain questions is likely also to have become less sure that God exists, and vice versa. They moreover suggest that this connection between the two changes at least partially independent of changes in political identification.

The results do not, however, determine any direction of causality between the two changes. The data set's only having two waves prevents direct inquiry into the temporal order of the changes. Hence the results are compatible with the theory presented above but do not specifically support it--for all we know, the miracle of faith springs into life in an individuals heart and then gives rise to an increase in concern for moral standards by opening the eyes, so to speak, of that individual to a reality that their previous sinful state had obscured from them.

Moreover, the results would be more significant if the data set contained a broader array of moral questions that were less politically charged. If such an array of questions were given to the respondents, it would 1) be possible to see if this relationship between moral saliency to practical questions and certainty in God holds more generally, and 2) be less of a concern that respondent's answers to the questions were influenced not just by their moral intuitions but by their partisan identification. If a 3+ wave survey containing such an array of questions were taken, the resulting analyses might be able to make evidence-based claims about the causal direction between these two variables.

The weight of the question of that direction cannot be underestimated. If the direction is from moral belief to religious belief, then faith itself would seem to depend not on a supernatural god but on other human beliefs and would offer no evidence for the existence of such a god. If the direction is from religious belief to moral belief, then there remains the possibility not only that a supernatural god exists but also that human reason is not sufficient for guiding human action. If justice is commanded by the one true God, the ancient political philosophers' failure to understand its coherence is only a consequence of their human limits.

## References
 
Baker, Joseph O'Brian, and Buster Smith. 2009. “None Too Simple: Examining Issues of Religious Nonbelief and Nonbelonging in the United States.” Journal for the Scientific Study of Religion 48(4): 719–33.
 
Bruell, Christopher. 1999. On the Socratic Education: an Introduction to the Shorter Platonic Dialogue. Rowman and Littlefield.
 
Campbell, Angus, Philip E. Converse, Warren E. Miller, and Donald E. Stokes. 1960. The American Voter. John Wiley.
 
Campbell, David, Geoffrey Layman, John Green, and Nathanael Sumaktoyo. 2018. “Putting Politics First: The Impact of Politics on American Religious and Secular Orientations.” American Journal of Political Science, 62: 551-565.
 
Djupe, Paul, Jacob Neiheisel, and Anand Sokhey. 2018. “Reconsidering the Role of Politics in Leaving Religion: The Importance of Affiliation.” American Journal of Political Science, 62: 161-175.
 
Hout, Michael, and Claude S. Fischer. 2002. “No Religious Preference: Politics and Generations.” American Sociological Review 67(2): 165–90.
 
Layman, Geoffrey C. 2001. The Great Divide: Religious and Cultural Conflict in American Party Politics. Columbia University Press.

Leibowitz, David. 2010. The Ironic Defense of Socrates: Plato's Apology. Cambridge University Press.
 
Patrikios, Stratos. 2008. “American Republican Religion?” Political Behavior 30(3): 367–89.
 
Putnam, Robert D., and David E. Campbell. 2010. American Grace: How Religion Divides and Unites Us. New York: Simon and Schuster.
 
Williams, Richard. 2006. "Generalized Ordered Logit/ Partial Proportional Odds Models for Ordinal Dependent Variables." The Stata Journal 6(1): 58-82.

\newpage

# Mathematical Appendix


## Introduction to the Ologit Model


The ologit model uses an ordinal dependent variable, $Y$. We assume that there is an unknown variable, $Y*$, that is a linear function of the observed independent variables and their unknown coefficients added to a random variable $\epsilon$ following the logistic distribution with mean 0 and variance 1:

$$Y*=\beta X + \epsilon$$

We then suppose that if the value of a particular $Y_{i}*$ is less than or equal to some $\mu_{0}$, then the value of $Y_{i}*$, $y_{i}*$, will equal $0$. If $y_{i}*$ is greater than $\mu_{0}$ and less than or equal to some $\mu_{1}$, then $y_{i}$ will equal $1$. In general, it is supposed that if $y_{i}*$ is greater than some $\mu_{k-1}$ and less than or equal to some $\mu_{k}$, then $y_{i}$ will be equal to $k$:

$$Y=0 \iff Y* \le \mu_{0}$$
$$...$$
$$Y=k \iff \mu_{k-1} < Y* \le \mu_{k}$$
$$...$$
$$Y=m \iff \mu_{m-1} < Y*$$

Since by supposition we do not know any $y_{i}*$, we have to estimate the $\mu$ parameters in addition to the $\beta$ parameters. In order to do this, we need to determine the probabilities for observing the different $y_{i}$ under this model.

We begin with $p_{0}=Pr_{Y_{i}}(y_{i}=0|x_{i})$. Since $$y_{i}=0 \iff y_{i}*\le \mu_{0},$$ we have $$Pr_{Y_{i}}(y_{i}=0|x_{i}) = Pr_{Y_{i}*}(y_{i}* \le \mu_{0}|x_{i}).$$ Since $$y_{i}*= x_{i}'\beta +\epsilon_{i},$$ $$Pr_{Y_{i}*}(y_{i}* \le \mu_{0}|x_{i}) = Pr( x_{i}'\beta +\epsilon \le \mu_{0}|x_{i}),$$ and this in turn equals $$Pr(\epsilon \le \mu_{0}- x_{i}'\beta |x_{i}).$$ Since the random variable $\epsilon$ follows the logistic distribution, this latter probability is equal to the CDF of the logistic distribution at $\mu_{0}- x_{i}'\beta .$ Hence $$p_{0} = Pr_{Y_{i}}(y_{i}=0|x_{i}) = \Lambda(\mu_{0}- x_{i}'\beta ).$$

What if $y_{i}$ is greater than $0$ but less than $m$? In that case, by reasoning analogous to that given above, 
$$p_{k}=Pr(Y_{i}=k|x_{i})=Pr(\mu_{k-1}<Y_{i}*\le\mu_{i}|x_{i})$$ 
$$=Pr(\mu_{k-1}< x_{i}'\beta +\epsilon \le \mu_{k}|x_{i})$$
$$=Pr(\mu_{k-1}- x_{i}'\beta <\epsilon\le \mu_{k}- x_{i}'\beta |x_{i})$$ 
$$=\Lambda(\mu_{k}- x_{i}'\beta )-\Lambda(\mu_{k-1} - x_{i}'\beta )$$

Finally, for the probability that $y_{i}=m$, we have 
$$p_{m}=Pr(Y_{i}=m|x_{i})=Pr(\mu_{m-1}<Y_{i}*|x_{i})$$ 
$$=Pr(\mu_{m-1}< x_{i}'\beta +\epsilon|x_{i})$$
$$=Pr((\mu_{m-1}- x_{i}'\beta <\epsilon)$$
$$=1-\Lambda(\mu_{m-1}- x_{i}'\beta )$$

If we let $\gamma_{i}^{0}$ equal $1$ if $Y_{i}$ equals $0$ and equal $1$ if $Y_{i}$ equals anything else, and likewise down through $\gamma_{i}^{m}$ equalling $1$ iff $Y_{i}$ equals $m$ and equalling $0$ otherwise, then the conditional PDF for $Y_{i}$ would be 
$$Pr_{Y_{i}}(y_{i}|\beta,\mu_{0},...,\mu_{m-1},x_{i})=p_{0}^{\gamma_{i}^{0}}p_{1}^{\gamma_{i}^{1}}...p_{m}^{\gamma_{i}^{m}}$$
and the conditional PMF for the whole sample would be 
$$Pr_{Y_{i}}(y_{1}, . . . ,y_{n}|\beta,\mu_{0},...,\mu_{m-1},x_{1},...,x_{n})=\prod_{i=0}^{n}p_{0}^{\gamma_{i}^{0}}p_{1}^{\gamma_{i}^{1}}...p_{m}^{\gamma_{i}^{m}}.$$ 
Changing the conditions, we get the likelihood function for the sample: 
$$L(\beta,\mu_{0},...,\mu_{m-1}|y_{1}, . . . ,y_{n},x_{1},...,x_{n})=\prod_{i=0}^{n}p_{0}^{\gamma_{i}^{0}}p_{1}^{\gamma_{i}^{1}}...p_{m}^{\gamma_{i}^{m}}.$$ 


## Introduction to MLE


But what do we need this likelihood function for? Enter maximum likelihood estimation. MLE is a method for estimating population parameters from sample data. In the model above (and in any other model with parameters that can be estimated by MLE) we have a measurable dependent variable, Y, that we assume depends for its value on the product of measurable independent variables, the vector X, and immeasurable population parameters, the vector B. More precisely, we assume that Y depends on these two in a probabilistic way, and for this reason we included the logistic random variable in our model above. Now, for any given values of our population parameters, we can determine the probability that we observe what we have observed for our dependent variables. The function that maps any given element in the set of all possible population parameter vectors to the probability of observing what we have observed if that element is the true population parameter vector is the likelihood function. That function is none other than the joint pdf of all our observations of our dependent variable conditional on certain parameters. Hence, if we were very poor political scientists and had made only one observation and were using a model in which the dependent variable of interest depends on three independent variables but nevertheless were also perfectly competent statisticians, we would write

$$L(\beta) = Pr(y|\beta),$$ where $\beta$ is a vector of the three parameters we want to estimate.

Once we have the likelihood function determined by our model, we maximize it--the parameter estimates given by MLE are those that have the highest likelihood of producing the values we observed for our dependent variables. To maximize a function, we first take its derivative, determine the elements of its domain for which the derivative is 0, and then select of the previous elements the one for which the second derivative of the function is negative, to ensure we have a maximum. In the unlikely event that there is more than a single local maximum, we would select the largest of them. 

Whenever we have made more than one observation, however, our likelihood function will be of the following form:

$$L(\beta) = \prod_{i=0}^{n}Pr(y_{i}|\beta).$$

This is because we assume that we are (and hopefully in fact are) dealing with independent variables, i.e. that the observed value of the dependent variable for any one of our observations does not affect the observed value of the dependent variable for any of our other observations. This being the case, the probability of our observing the values we observed will be the product of the probability of observing each value given certain parameters. In this way we come to the equation above. But taking the derivative of such a function would be undesirable. If, however, we take the natural log of the function, which we then call the log-likelihood function, we would only have to deal with a sum of probability functions, since $ln(f(x)g(x))=log(f(x))+log(g(x))$. But will the maximum of the log likelihood function always be a maximum of the likelihood function? The answer turns out to be yes, as the following equation indicates.

$$\frac{d(ln(L(\beta)))}{d\beta}=\frac{1}{L(\beta)}\frac{d(L(\beta))}{d\beta}$$

because the derivative on the left side will be $0$ only when $\frac{d(L(\beta))}{d\beta}$ is $0$, and because whenever $\frac{d(L(\beta))}{d\beta}$ is $0$, the derivative on the left side will be $0$, we can affirm that the extrema of the log-likelihood function are the same as those of the likelihood function. To show that the extrema of the log-likelihood function are the same type of extrema as those of the likelihood function, i.e., in particular, that the maxima of the one are the maxima of the other, the following observation suffices. The second derivative of the log-likelihood function is the following:

$$\frac{d^{2}ln(L(\beta))}{d\beta^{2}}=\frac{d\frac{1}{L(\beta))}}{d\beta}\frac{d(L(\beta))}{d\beta}+ \frac{1}{L(\beta)}\frac{d\frac{d(L(\beta))}{d\beta}}{d\beta}.$$
$$=-(\frac{d(L(\beta))}{d\beta}\frac{1}{L(\beta)})^2+\frac{d^2(L(\beta))}{d\beta^2}\frac{1}{L(\beta)}$$

Since the squared term will always be $0$ by the assumption that we have an extremum, we need only examine the second term. Since $L(\beta)$ is always positive, $\frac{1}{L(\beta)}$ is always positive. The sign of this expression when $\beta$ is an extremum thus depends entirely on the sign of the second derivative of $L(\beta)$ and $\beta$ will be a maximum of the log-likelihood function iff it is a maximum of the likelihood function.


## Maximum Likelihood Estimation of the Ologit Model


Having seen all this, we know how to estimate the $\mu$ and $\beta$ parameters for our ologit model. We first take the log of the likelihood function of the entire sample: 

$$ln(L(\beta,\mu_{0},...,\mu_{m-1}|y_{1}, . . . ,y_{n},x_{1},...,x_{n}))=ln(\prod_{i=0}^{n}p_{0}^{\gamma_{i}^{0}}p_{1}^{\gamma_{i}^{1}}...p_{m}^{\gamma_{i}^{m}})$$ 
$$=\sum_{i=0}^{n}ln(p_{0}^{\gamma_{i}^{0}}p_{1}^{\gamma_{i}^{1}}...p_{m}^{\gamma_{i}^{m}}),$$
and then maximize the function with respect to the $\mu$ and $\beta$ parameters, treating the observed $y_{i}$ and $x_{i}$ as constants. 


## The Gologit Model


Now for the generalized ologit model: I will first provide the model and then explain its relation to the ologit model, the motivation for it, and how its parameters can be estimated. The gologit model is as follows:

$$Pr(Y_{i}>j|x_{i})=\frac{exp(x_{i}'\beta_{j} -\mu_{j})}{1+exp(x_{i}'\beta_{j}-\mu_{j})}, j=1,2,...,m-1$$
  *note*: I reverse the sign on William's alphas and replace them with mu's to keep notation consistent with my presentation of the ologit model above. See Williams, slide 13 of https://www3.nd.edu/~rwilliam/stats3/Gologit2Part1.pdf

The first difference between this model and the ologit model that strikes one is that the probability on the left side of the equation is not the probability that the value of an observed dependent variable equals something but the probability that it is greater than something. So let us first show that the ologit model can be represented in this way. 

For any discrete random variable $Y$, $$Pr(Y=j+1) = Pr(Y>j) - Pr(Y>j+1)$$. Knowing this, we can re-represent our ologit model with probabilities that $Y$ is greater than some $j$. According to our ologit model,

$$Pr(Y_{i}=0|x_{i}) = \Lambda(\mu_{0}- x_{i}'\beta )$$

Given what we have just seen, we know that 

$$Pr(Y_{i}=0|x_{i}) = 1 - Pr(Y_{i}>0|x_{i}) = \Lambda(\mu_{0}- x_{i}'\beta )$$

Now for algebra:

$$1 - Pr(Y_{i}>0|x_{i}) = \Lambda(\mu_{0}- x_{i}'\beta )$$

$$Pr(Y_{i}>0|x_{i}) = 1-\Lambda(\mu_{0}- x_{i}'\beta )$$

$$=1-\frac{exp(\mu_{0}- x_{i}'\beta )}{1+exp(\mu_{0}- x_{i}'\beta )}$$
$$=\frac{1}{1+exp(\mu_{0}- x_{i}'\beta )}$$
$$=\frac{exp( x_{i}'\beta )}{exp( x_{i}'\beta )+exp(\mu_{0})}$$
$$Pr(Y_{i}>0|x_{i})=\frac{exp( x_{i}'\beta -\mu_{0})}{1+exp( x_{i}'\beta -\mu_{0})}$$
Now we turn to the probability that $Y_{i}=1$:
$$Pr(Y_{i}=1|x_{i})=Pr(Y_{i}>0|x_{i})-Pr(Y_{i}>1|x_{i})$$
$$Pr(Y_{i}>1|x_{i})=Pr(Y_{i}>0|x_{i})-Pr(Y_{i}=1|x_{i})$$
$$=1-\Lambda(\mu_{0}- x_{i}'\beta )-(\Lambda(\mu_{1}- x_{i}'\beta )-\Lambda(\mu_{0}- x_{i}'\beta ))$$
$$=1-\Lambda(\mu_{1}- x_{i}'\beta ).$$
By following the same procedure as above, this term can be algebraically transformed so that we have
$$Pr(Y_{i}>1|x_{i})=\frac{exp( x_{i}'\beta -\mu_{1})}{1+exp( x_{i}'\beta -\mu_{1})}.$$
We could continue this process until representing our entire logit model in terms of these probabilities that $Y_{i}$ is greater than some $y_{i}$. If we did finish this process, the final equation would be
$$Pr(Y_{i}>m-1|x_{i})=\frac{exp( x_{i}'\beta -\mu_{m-1})}{1+exp( x_{i}'\beta -\mu_{m-1})},$$
and we could write all of this more compactly like this:
$$Pr(Y_{i}>j|x_{i})=\frac{exp( x_{i}'\beta -\mu_{j})}{1+exp( x_{i}'\beta -\mu_{j})}, j=1,2,...,m-1$$
Our ologit model thus represented is nearly the same as the gologit model written above, but there is one important difference: the ologit model does not have subscripts under the $\beta$ parameters. This means that the ologit model assumes that the same $\beta$ parameters apply no matter whether we are determining the probability that $Y_{i}>0$ or $Y_{i}>1$, whereas the gologit model allows $\beta$ to vary depending on $j$. In other words, the ologit model has the built in assumption that the effect of an independent variable on the probability that our dependent variable is greater than some $j$ is the same effect it has on the probability that our dependent variable value is greater than $j+1$. This is the proportional odds assumption. But as we saw in the paper above, this assumption does not always hold across all the variables in the model, and for these cases, it is safer to use the gologit model, which allows one to estimate different parameters for a single variable. Ologit is in this way the special case of the gologit model when all parameters are restricted to reflect the proportional odds assumption, or, to put this another way, we arrive at the gologit model by removing the restriction on the $\beta$ parameter from the ologit model. 


### Interlude: Interpretation of Ologit Coefficients


The proportional odds assumption we have just seen is connected to a standard interpretation of the coefficients of an ologit model. Because the ologit model has the proportional odds assumption baked into it, it can easily be shown, using the statements above, that for any $j$ that is on the support of the PMF of $Y_{i}$ but not the greatest element of that support,

$$\frac{Pr(Y_{i}>j|x_{i})}{Pr(Y_{i}\le j|x_{i})}=exp( x_{i}'\beta -\mu_{j}).$$
This means that an increase of 1 unit in any $x_{ib}$, meaning the $b^{th}$ element of the vector $x_{i}$, or, equivalently, the observed value of the $b^{th}$ independent variable for observation $i$, will increase the odds that $Y_{i}>j$ by a factor of $exp(\beta_{b})$, where $\beta_{b}$ is the coefficient for the $b^{th}$ element of the vector$x_{i}$. 

This interpretation of the coefficients of the ologit model works just as well for any coefficients in a gologit model that are constrained. For those coefficients of a gologit model that are not constrained, the simplest interpretation is that a one unit increase in their corresponding independent variables will increase the odds that $Y_{i}>j$ by a factor of the exponential of themselves, but only for the $j$ they are restricted to. To give an intuitive example, if our dependent variable were number of alcoholic drinks per month, a variable that might break the proportional odds assumption and need unrestricted coefficients is moderation. Since the moderate person seeks to have neither too little or too much, an increase in moderation could lead to an increase in the odds that the number of drinks per month is greater than $0$, but the same increase in moderation could lead to a massive *decrease* in the odds that the number of drinks per month is greater than $200$.


## MLE of the Gologit Model


To estimate the parameters of a gologit model, we need again to find the likelihood function, but most of our work has already been done for us. We saw above that the likelihood function for the ologit model was 

$$L(\beta,\mu_{0},...,\mu_{m-1})=\prod_{i=0}^{n}p_{0}^{\gamma_{i}^{0}}p_{1}^{\gamma_{i}^{1}}...p_{m}^{\gamma_{i}^{m}}.$$ 
Since $p_{j}=Pr(Y_{i}=j),$ and since we know that $Pr(Y_{i}=j)=Pr(Y_{i}>j-1)-Pr(Y_{i}>j),$ the gologit likelihood function will be
$$L(\beta_{0},...,\beta_{m-1},\mu_{0},...,\mu_{m-1})=\prod_{i=0}^{n}(1-Pr(Y_{i}>0))^{\gamma_{i}^{0}}(Pr(Y_{i}>0)-Pr(Y_{i}>1))^{\gamma_{i}^{1}}...(Pr(Y_{i}>m-1))^{\gamma_{i}^{m}},$$ 
and one need only maximize it with repect to the parameters to find the estimates.